\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2022-03-25 doc/USER_MANUAL/03_mesh_generation.tex
updates doc

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models.
Note that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation, set \texttt{Qkappa} to 9999 here.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.

\item For tomographic velocity models, please read Chapter \ref{cha:-Changing-the}
and Section \ref{sec:Using-tomographic} `Using external tomographic
Earth models' for further details.
\end{itemize}
\item [{nodes\_coords\_file}] Contains the point locations in Cartesian
coordinates of the mesh element corners.
\item [{mesh\_file}] Contains the mesh element connectivity. The hexahedral
elements can have 8 or 27 nodes.\\
 See picture doc/mesh\_numbering\_convention/numbering\_convention\_27\_nodes.jpg
to see\\
 in which (standard) order the points must be cited. In the case of
8 nodes, just include the first 8 points.
\item [{free\_or\_absorbing\_surface\_file\_zmax}] Contains the free surface
connectivity or \\
 the surface connectivity of the absorbing boundary surface at the
top (Zmax), \\
 depending on whether the top surface is defined as free or absorbing
(\texttt{STACEY\_INSTEAD\_OF\_FREE\_SURFACE} in \texttt{DATA/Par\_file}). \\
 You should put both the surface of acoustic regions and of elastic
regions in that file; that is, list all the element faces that constitute
the surface of the model in that file.
\item [{absorbing\_surface\_file\_xmax}] Contains the surface connectivity
of the absorbing boundary surface at Xmax \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_xmin}] Contains the surface connectivity
of the absorbing boundary surface at Xmin \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_ymax}] Contains the surface connectivity
of the absorbing boundary surface at Ymax \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_ymin}] Contains the surface connectivity
of the absorbing boundary surface at Ymin \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_bottom}] Contains the surface connectivity
of the absorbing boundary surface at the bottom (Zmin) \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\end{description}
These mesh files are needed as input files for the partitioner \texttt{xdecompose\_mesh}
to load-balance the mesh. Please see the next section for further
details.

@@

An efficient partitioning is important because it leverages the overall
running time of the application. It amounts to balance the number
of elements in each slice while minimizing the communication costs
resulting from the placement of adjacent elements on different processors.
\texttt{decompose\_mesh} depends on the SCOTCH library \citep{PeRo96},
which provides efficient static mapping, graph and mesh partitioning
routines. SCOTCH is a free software package developed by Fran\c{c}ois
Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable
from the web page \url{https://gforge.inria.fr/projects/scotch/}.\\

@@

The mesher generates the necessary partitions in parallel, one set
for each of the $\nprocxi\times\nproceta$ slices that constitutes
the mesh (see Figure~\ref{fig:For-parallel-computing}). After the
mesher finishes, you can log in to one of the compute nodes and view
the contents of the \texttt{LOCAL\_PATH} directory to see the files
generated by the mesher. These files will be needed for creating the
distributed databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{Par\_file}, e.g. in directory \texttt{OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
%
\item [{\texttt{NMATERIALS}}] The number of different materials in your
model. In the following lines, each material needs to be defined as:

\begin{verbatim}
material_ID rho vp vs Q anisotropy_flag domain_ID
\end{verbatim}

\noindent
where
\begin{itemize}
\item \texttt{Q} : quality factor (0=no attenuation) for shear attenuation $Q_{\mu}$
\item \texttt{anisotropy\_flag} : 0=no anisotropy / 1,2,.. check with implementation
in \texttt{aniso\_model.f90}
\item \texttt{domain\_id} : 1=acoustic / 2=elastic
\end{itemize}
\item [{\texttt{NREGIONS}}] The number of regions in the mesh. In the following
lines, because the mesh is regular or 'almost regular', each region
is defined as:
\begin{verbatim}
NEX_XI_BEGIN NEX_XI_END NEX_ETA_BEGIN NEX_ETA_END NZ_BEGIN NZ_END material_ID
\end{verbatim}
\end{description}
%
The \texttt{INTERFACES\_FILE} parameter of \texttt{Mesh\_Par\_File}
defines the file which contains the settings of the topography grid
and of the interfaces grids. Topography is defined as a set of elevation
values on a regular 2D grid. It is also possible to define interfaces
between the layers of the model in the same way. The file needs to
define several parameters:
\begin{itemize}
\item The number of interfaces, including the topography. This needs to
be set at the first line. Then, from the bottom to the top of the
model, you need to define the grids with:
\item \texttt{SUPPRESS\_UTM\_PROJECTION} flag as described previously,
\item number of points along $x$ and $y$ direction (NXI and NETA),
\item minimal $x$ and $y$ coordinates (LONG\_MIN and LAT\_MIN),
\item spacing between points along $x$ and $y$ (SPACING\_XI and SPACING\_ETA)
and
\item the name of the file which contains the elevation values (in $y$.$x$
increasing order).
\end{itemize}
%
At the end of this file, you simply need to set the number of spectral
elements in the vertical direction for each layer. We provide a few
models in the {\texttt{EXAMPLES/}} directory.\\

Now that you have set the appropriate parameters in the \texttt{Mesh\_Par\_file}
and have compiled the mesher, you are ready to run it!
\footnote{
Depending on your system, please check how to run executables with MPI.
To run the mesher on your cluster, we provide some example scripts for you to modify accordingly in \texttt{utils/Cluster/}.
For some installations, you might need to provide a file
that tells MPI what compute nodes to use for the simulations. In this case, the
file must have a number of entries (one entry per line) at least equal
to the number of processors needed for the run.
A sample file name is \texttt{mymachines}. This file is not used by the mesher
or solver, but is required by the \texttt{go\_mesher***} and \texttt{go\_solver**}
default job submission scripts provided in directory \texttt{utils/Cluster/}.
To run the mesher, it is most easily accomplished based upon the \texttt{go\_mesher***} script.
When you run on a PC cluster, the script assumes that the nodes are
named n001, n002, etc. If this is not the case, change the \texttt{tr
-d `n'} line in the script. You may also need to edit the last command
at the end of the script that invokes the \texttt{mpirun} command.
See Chapter~\ref{cha:Scheduler}
for information about running the code on a system with a scheduler,
e.g., LSF.
}
Mesher output is provided in the \texttt{OUTPUT\_FILES/} directory
in \texttt{output\_mesher.txt}; this file provides lots of details
about the mesh that was generated.\\

@@

>>>>

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models.
Note that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation, set \texttt{Qkappa} to 9999 here.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.

To convert one to the other see \texttt{doc/note\_on\_Qkappa\_versus\_Qp.pdf} and the helper code\\
\texttt{conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90} in folder
\texttt{utils/attenuation/}.

\item For tomographic velocity models, please read Chapter \ref{cha:-Changing-the}
and Section \ref{sec:Using-tomographic} `Using external tomographic
Earth models' for further details.
\end{itemize}
\item [{nodes\_coords\_file}] Contains the point locations in Cartesian
coordinates of the mesh element corners.
\item [{mesh\_file}] Contains the mesh element connectivity. The hexahedral
elements can have 8 or 27 nodes.\\
 See picture \texttt{doc/mesh\_numbering\_convention/numbering\_convention\_27\_nodes.jpg}
to see\\
 in which (standard) order the points must be cited. In the case of
8 nodes, just include the first 8 points.
\item [{free\_or\_absorbing\_surface\_file\_zmax}] Contains the free surface
connectivity or \\
 the surface connectivity of the absorbing boundary surface at the
top (Zmax), \\
 depending on whether the top surface is defined as free or absorbing
(\texttt{STACEY\_INSTEAD\_OF\_FREE\_SURFACE} in \texttt{DATA/Par\_file}). \\
 You should put both the surface of acoustic regions and of elastic
regions in that file; that is, list all the element faces that constitute
the surface of the model in that file.
\item [{absorbing\_surface\_file\_xmax}] Contains the surface connectivity
of the absorbing boundary surface at Xmax \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_xmin}] Contains the surface connectivity
of the absorbing boundary surface at Xmin \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_ymax}] Contains the surface connectivity
of the absorbing boundary surface at Ymax \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_ymin}] Contains the surface connectivity
of the absorbing boundary surface at Ymin \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\item [{absorbing\_surface\_file\_bottom}] Contains the surface connectivity
of the absorbing boundary surface at the bottom (Zmin) \\
(also needed in the case of C-PML absorbing conditions, in order for
the code to be able to impose Dirichlet conditions on their outer
edge).
\end{description}
These mesh files are needed as input files for the partitioner \texttt{xdecompose\_mesh}
to load-balance the mesh. Please see the next section for further
details.

@@

An efficient partitioning is important because it leverages the overall
running time of the application. It amounts to balance the number
of elements in each slice while minimizing the communication costs
resulting from the placement of adjacent elements on different processors.
\texttt{decompose\_mesh} depends on the SCOTCH library \citep{PeRo96},
which provides efficient static mapping, graph and mesh partitioning
routines. SCOTCH is a free software package developed by Fran\c{c}ois
Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable
from the web page \url{https://gitlab.inria.fr/scotch/scotch}.\\

@@

The mesher generates the necessary partitions in parallel, one set
for each of the $\nprocxi\times\nproceta$ slices that constitutes
the mesh (see Figure~\ref{fig:For-parallel-computing}). After the
mesher finishes, you can log in to one of the compute nodes and view
the contents of the \texttt{LOCAL\_PATH} directory to see the files
generated by the mesher. These files will be needed for creating the
distributed databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{Par\_file}, e.g. in directory \texttt{OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
%
\item [{\texttt{NMATERIALS}}] The number of different materials in your
model. In the following lines, each material needs to be defined as:
%
\begin{itemize}[leftmargin=-8ex,itemindent=8ex]
\item[] acoustic/elastic material
{\small
\begin{verbatim}
material_ID rho vp vs Qkappa Qmu anisotropy_flag domain_ID
\end{verbatim}
}
\item[] poroelastic material
{\small
\begin{verbatim}
material_id rho_s rho_f phi tort kxx kxy kxz kyy kyz kzz kappa_s kappa_f kappa_fr eta mu_fr domain_id
\end{verbatim}
}
\item[] tomographic
{\small
\begin{verbatim}
material_id type-keyword domain-name tomo-filename tomo_id domain_id
\end{verbatim}
}
\end{itemize}
For acoustic and elastic material, you would specify
\begin{itemize}
\item \texttt{material\_ID} : a material ID number (1,2,..) to identify the material for the following mesh region selection.
\item \texttt{rho}, \texttt{vp}, \texttt{vs} : density (in $kg/m^3$), $V_p$ and $V_s$ (in $m/s$) wave speeds
\item \texttt{Qkappa, Qmu} : quality factor (0=no attenuation) for bulk $Q_{\kappa}$ and shear attenuation $Q_{\mu}$
\item \texttt{anisotropy\_flag} : 0=no anisotropy / 1,2,.. check with implementation
in \texttt{aniso\_model.f90}
\item \texttt{domain\_id} : 1=acoustic / 2=elastic / 3=poroelastic
\end{itemize}
\noindent
For poroelastic materials, you would specify the corresponding poroelastic properties.
Please read the section below for more on poroelastic parameters.\\

\noindent
For tomographic models, you would need to specify a negative \texttt{material\_id} (-1,-2,..)
and a corresponding \texttt{domain\_id} for either acoustic (1) or elastic (2) domains. For example a line like:
{\small
\begin{verbatim}
-1 tomography elastic tomography_model.xyz 0 2
\end{verbatim}
}
where the specifications \texttt{tomography}, \texttt{tomography\_model.xyz}, \texttt{0} are always used by default and
the keyword \texttt{elastic} or \texttt{acoustic} will be defined internally according to your specified \texttt{domain\_id}.
%
\item [{\texttt{NREGIONS}}] The number of regions in the mesh. In the following
lines, because the mesh is regular or 'almost regular', each region
is defined as:
{\small
\begin{verbatim}
NEX_XI_BEGIN NEX_XI_END NEX_ETA_BEGIN NEX_ETA_END NZ_BEGIN NZ_END material_ID
\end{verbatim}
}
\end{description}
%
The \texttt{INTERFACES\_FILE} parameter of \texttt{Mesh\_Par\_File}
defines the file which contains the settings of the topography grid
and of the interfaces grids. Topography is defined as a set of elevation
values on a regular 2D grid. It is also possible to define interfaces
between the layers of the model in the same way. The file needs to
define several parameters:
\begin{itemize}
\item The number of interfaces, including the topography. This needs to
be set at the first line. Then, from the bottom to the top of the
model, you need to define the grids with:
\item \texttt{SUPPRESS\_UTM\_PROJECTION} flag as described previously,
\item number of points along $x$ and $y$ direction (NXI and NETA),
\item minimal $x$ and $y$ coordinates (LONG\_MIN and LAT\_MIN),
\item spacing between points along $x$ and $y$ (SPACING\_XI and SPACING\_ETA)
and
\item the name of the file which contains the elevation values (in $y$.$x$
increasing order).
\end{itemize}
%
At the end of this file, you simply need to set the number of spectral
elements in the vertical direction for each layer. We provide a few
models in the {\texttt{EXAMPLES/}} directory.\\

Now that you have set the appropriate parameters in the \texttt{Mesh\_Par\_file}
and have compiled the mesher, you are ready to run it!\\

Depending on your system, please check how to run executables with MPI.
To run the mesher on your cluster, we provide some example scripts for you to modify accordingly in \texttt{utils/Cluster/}.
For some installations, you might need to provide a file
that tells MPI what compute nodes to use for the simulations. In this case, the
file must have a number of entries (one entry per line) at least equal
to the number of processors needed for the run.
A sample file name is \texttt{mymachines}. This file is not used by the mesher
or solver, but is required by the \texttt{go\_mesher***} and \texttt{go\_solver**}
default job submission scripts provided in directory \texttt{utils/Cluster/}.
To run the mesher, it is most easily accomplished based upon the \texttt{go\_mesher***} script.
When you run on a PC cluster, the script assumes that the nodes are
named \texttt{n001}, \texttt{n002}, etc.
If this is not the case, change the \texttt{tr -d `n'} line in the script.
You may also need to edit the last command
at the end of the script that invokes the \texttt{mpirun} command.
See Chapter~\ref{cha:Scheduler}
for information about running the code on a system with a scheduler,
e.g., LSF.\\

Mesher output is provided in the \texttt{OUTPUT\_FILES/} directory
in \texttt{output\_mesher.txt}; this file provides lots of details
about the mesh that was generated.\\

@@

\section{Poroelastic materials}\label{cha:Poroelastic-materials}

The present version of SPECFEM can handle fully saturated porous simulations.
At the moment, the code implements Biot's equation.
However, the code cannot calculate partially saturated cases in its current state.\\

The way we prescribe material property for porous material in SPECFEM3D depends on wheter the internal mesher \texttt{xmeshfem3D}
or an external mesher like CUBIT is taken.
For external meshers which provide the resulting meshes in a folder like \texttt{MESH/},
we use a file \texttt{nummaterial\_poroelastic\_file}, which needs to be added to this directory \texttt{MESH/}.
The line format for poroelastic materials is as follow:
{\small \begin{verbatim}
  rhos rhof phi c kxx kxy kxz kyy kyz kzz Ks Kf Kfr etaf mufr
\end{verbatim}}
\begin{description}
\item where\\
\texttt{rho\_s} = solid density,\\
\texttt{rho\_f} = fluid density,\\
\texttt{phi} = porosity,\\
\texttt{tort} = tortuosity,\\
\texttt{kxx} = xx component of permeability tensor,\\
\texttt{kxy} = xy,yx components of permeability tensor,\\
\texttt{kyy} = yy component of permeability tensor,\\
\texttt{kxz} = xz,zx components of permeability tensor,\\
\texttt{kzz} = zz component of permeability tensor,\\
\texttt{kappa\_s} = solid bulk modulus,\\
\texttt{kappa\_f} = fluid bulk modulus,\\
\texttt{kappa\_fr} = frame bulk modulus,\\
\texttt{eta\_f} = fluid viscosity,\\
\texttt{mu\_fr} = frame shear modulus.\\
\end{description}
Using an external mesh (for instance coming from CUBIT/TRELIS), poroelastic materials have the ID number 3, while 1 is acoustic and 2 is elastic (see the example in the package: EXAMPLES/homogeneous\_poroelastic).\\

\noindent
In case the internal mesher \texttt{xmeshfem3D} is used, the same poroelastic material properties will be defined in the \texttt{Mesh\_Par\_file}.
See the above section~\ref{cha:Running-the-Mesher-Meshfem3D} on how the corresponding line format would look like.\\

\noindent
Note that attenuation is not implemented yet for poroelastic domains, thus there is no need for specifiying Q values.
